Space for WHAM: a multi-region, multi-stock generalization of the Woods Hole Assessment Model with an application to black sea bass

### 2025-07-24 paper/paper_accepted_crossref_supp.tex
separate supp materials

\externaldocument{supp_final}

@@

State-space age-structured stock assessment models can be used to estimate time and age-varying population attributes as random effects using maximum marginal likelihood or Bayesian fitting procedures (Nielsen and Berg 2014; Cadigan 2016; Miller et al. 2016b). This estimation approach is considered an essential feature of gold-standard assessment models that we use in tactical management of commercially important fish stocks (Punt et al. 2020). The State-space Assessment Model (SAM, Nielsen and Berg 2014) continues to be developed and remains widely used within the International Council for the Exploration of the Sea (ICES) to assess European fish stocks. Various state-space models are being used to manage cod and plaice stocks in the waters of Eastern Canada (Perreault et al. 2020; Varkey et al. 2022) and to the south, the Woods Hole Assessment Model (WHAM, Stock and Miller 2021) is now used to assess many fish stocks in the Northwest Atlantic Ocean (NEFSC 2022a, 2022b; NEFSC 2024a).

WHAM is an R package developed and maintained by scientists at NOAA's Northeast Fisheries Science Center (\url{https://timjmiller.github.io/wham}). WHAM can be configured to fit a wide range of age-structured models from traditional statistical catch-at-age models without any random effects, to models with several time and age varying process errors and effects of environmental covariates on various population parameters. Like SAM, WHAM models are built using the Template Model Builder package (TMB, Kristensen et al. 2016) which provides a computationally efficient means of fitting an extremely wide class of models with random effects. WHAM has undergone active development since its creation and includes random effects options for recruitment, inter-annual transitions of numbers at age (hereafter referred to as ``apparent survival''), fishery and index selectivity, natural mortality, and catchability.

However, WHAM has only allowed consideration of a single stock and region, and no seasonal changes in stock dynamics. Using such models for stocks that have subcomponents with varying seasonal movement can provide incorrect inferences and poor management advice (Ying et al. 2011; Cao et al. 2014; Bosley et al. 2022). Furthermore, the ability to account for spatial structure and model multiple stocks are also important features of leading-edge assessment modeling frameworks (Punt et al. 2020). We describe here the implementation of these features and other extensions since Stock and Miller (2021) in WHAM version 2.0. Many of these new configuration options can also be useful when modeling a single stock and region.

We developed this extension of WHAM in concert with a research track assessment process for black sea bass (BSB, \textit{Centropristis striata}; NEFSC 2023), which is a high profile stock in the Northeast U.S. and important to both commercial fishing fleets and recreational anglers. BSB are targeted from Maine to North Carolina in both state and federal waters. Past stock assessments have estimated that the stock is healthy (not overfished and overfishing not occurring). Those assessments assumed two stock components (north and south), each assessed with a separate statistical catch at age model, with results subsequently combined to inform management. However, the assessments exhibited strong retrospective patterns. The stock was split into two regional components to account for spatial dynamics and improve model diagnostics (ASMFC 2016). Additionally, BSB have complex spatial dynamics and make seasonal migrations, moving inshore in the spring and offshore in the fall (Moser and Shepherd 2009). BSB distribution has been linked to warming waters on the northwest shelf (Bell et al. 2015) and it is hypothesized that the distribution and productivity of the BSB stock is especially susceptible to long-term changes in temperature causing increases in abundance further north (Hare et al. 2016). Thus, a focus of the BSB research track assessment was to develop an assessment platform that could capture multiple stock components, complex seasonal movement, and explore environmental drivers of population dynamics (NEFSC 2023). We denote the fish in the north as separate stock components rather than stocks because both components are managed as a single stock, but recognizing differences in migration between fish originating in the two areas.

To illustrate the usage of WHAM 2.0 we apply it to the two stock components of BSB off the coast of the Northeast United States (NEUS) using most of the model assumptions that were accepted during the peer-review process (NEFSC 2023). However, here we evaluate evidence for alternative hypotheses of temporal variation and effects of bottom temperature, specifically, on recruitment and natural mortality of age 1 individuals.

When modeling multiple stock components and regions, further dimensionality is added to the class of possible equilbrium reference points (e.g., Kapur et al. 2021). Short term projections are also an important part of the management process and the separation of observation and process errors in state-space models allows uncertainty in projected attributes such as recruitment and spawning stock biomass (SSB) without external simulation exercises. Finally, the inclusion of explicit effects of environmental covariates in the model also requires assumptions on how they are treated in projections and reference points. We demonstrate some of the options of WHAM version 2.0 to provide biological reference points and short-term projections.

@@

Many of the options and equations of WHAM version 2.0 are the same as those in Stock and Miller (2021), so we will only describe extensions and differences that have occurred since their first description of WHAM. The new version of WHAM can model multiple stocks, each with their own movement, harvest, and natural mortality. Much of the description below is for a specific stock \(s\), but, for simplicity, this subscript is implicit except when necessary. We have provided an Appendix A Table \ref{symbols} containing definitions of all mathematical notation in this paper.

@@

Because individuals may be alive in one of several regions or harvested by one of several fleets, it is helpful to consider these as distinct categories or states and treat the number of individuals occurring in each category over time as a multi-state model. Approaches to modeling transitions among these states may treat time discretely (e.g., Arnason 1972; Schwarz et al. 1993) or continuously (e.g., Hearn et al. 1987; Commenges 1999; Andersen and Keiding 2002). We can define a probability transition matrix (PTM) that describes the probability of individuals occurring in different states at the end of a time interval with duration \(\delta\) (columns), conditional on being in each of those states at the beginning of the interval (rows). For fish populations, these states would be defined as being alive in a particular region or deceased due to fishing from a particular fleet or natural mortality. The time interval \(i\) with duration \(\delta_i\) would be a season and the PTMs would be uniquely defined for each stock by year \(y\), age \(a\), and season \(i\). Each row and column of the PTM correspond to one of the states: alive in region \(r\), dead in fleet \(f\), or dead from natural causes. The probabilities in each row sum to one (must be in one of the states) and assume an individual is in the state that corresponds to that row at the beginning of the interval. Given \(n_R\) regions and \(n_F\) fleets, the square PTM (\(n_R + n_F + 1\) rows and columns) as a function of sub-matrices is
\begin{equation}\label{eq:ptm}
  \mathbf{P}_{y,a,i} = \begin{bmatrix}
    \mathbf{O}_{y,a,i} & \mathbf{H}_{y,a,i} & \mathbf{D}_{y,a,i} \\
    0 & \mathbf{I}_{H} & 0\\
    0 & 0 & 1
  \end{bmatrix},
\end{equation}
where
\begin{equation*}
  \mathbf{O}_{y,a,i} = 
  \begin{bmatrix}
    O_{y,a,i}(1,1) & \cdots & O_{y,a,i}(1,n_R) \\
    \vdots & \ddots & \vdots \\
    O_{y,a,i}(n_R,1) & \cdots & O_{y,a,i}(n_R,n_R)
  \end{bmatrix}
\end{equation*}
is the \(n_R \times n_R\) matrix defining survival and occurring in each region at the end of the interval,
\begin{equation*} 
  \mathbf{H}_{y,a,i} = 
  \begin{bmatrix}
    H_{y,a,i}(1,1) & \cdots & H_{y,a,i}(1,n_F) \\
    \vdots & \ddots & \vdots \\
    H_{y,a,i}(n_R,1) & \cdots & H_{y,a,i}(n_R,n_F)
  \end{bmatrix}
\end{equation*}
is the \(n_R \times n_F\) matrix defining probabilities of being captured in each fleet during the interval, and \(\mathbf{D}_{y,a,i}\) is the \(n_R \times 1\) matrix of probabilities of dying due to natural mortality during the interval. We have the \(n_F\times n_F\) identity matrix \(\mathbf{I}_{H}\) for the states of capture by each fleet and a 1 for the natural mortality state because the probabilities of being in one of the mortality states given starting the interval in that state is one (no zombies allowed).

@@

When there is more than one region, WHAM can model survival and movement as processes occurring sequentially or simultaneously. The sequential assumption is used widely in spatially explicit models (e.g., Stock Synthesis, Methot and Wetzel 2013). Under the sequential assumption, survival and death occur over the interval and movement among regions occurs instantly at either the beginning or the end of the interval. WHAM is configured to have movement occur after survival:
\begin{equation*}
  \mathbf{O}_{y,a,i} = \mathbf{S}_{y,a,i}\boldsymbol{\mu}_{y,a,i},
\end{equation*}
where \(\mathbf{S}_{y,a,i}\) is a \(n_R \times n_R\) diagonal matrix of proportions surviving in each region (given they start in that region)
\begin{equation*}
\mathbf{S}_{y,a,i} = 
  \begin{bmatrix}
    e^{-Z_{y,a,i,1}\delta_i} & 0 & \cdots & 0 \\
    0 & e^{-Z_{y,a,i,2}\delta_i} & \cdots & 0 \\
    \vdots & \vdots & \ddots & \vdots \\
    0 & \cdots & 0 & e^{-Z_{y,a,i,n_R}\delta_i}
  \end{bmatrix}
\end{equation*}
and \(\boldsymbol{\mu}_{y,a,i}\) is a \(n_R \times n_R\) matrix of proportions moving from one region to another or staying in the region they occurred at the beginning of the interval:
\begin{equation*}
\boldsymbol{\mu}_{y,a,i} = 
  \begin{bmatrix}
    1-\sum_{r' \neq 1} \mu_{1\rightarrow r',y,a,i} & \mu_{1\rightarrow 2,y,a,i} & \cdots & \mu_{1\rightarrow R,y,a,i} \\
    \mu_{2\rightarrow 1,y,a,i} & 1-\sum_{r' \neq 2} \mu_{2\rightarrow r',y,a,i} & \cdots & \mu_{2\rightarrow R,y,a,i} \\
    \vdots & \vdots & \ddots & \vdots \\
    \mu_{R\rightarrow 1,y,a,i} & \cdots & \mu_{R\rightarrow R-1,y,a,i} & 1-\sum_{r' \neq R} \mu_{R\rightarrow r',y,a,i}
  \end{bmatrix}
\end{equation*}

@@

As in previous versions of WHAM, the apparent survival (and movement) from one year to another after recruitment can be treated deterministically or as functions of random effects. The predicted numbers at age \(a\) in year \(y\) for a given stock are vector analogs (\(\mathbf{N}_{O,y,a}\)) of the equations for numbers at age in the standard WHAM model (Stock and Miller 2021). For ages \(a = 2,\ldots, A-1\), where \(A\) is the plus group, the expected number alive in each region at the beginning of the following year and next age class age can be obtained from the first \(n_R\) elements of the vector
\[\mathbf{P}_{y-1,a-1}^T \mathbf{N}_{y-1,a-1}.\]
The numbers alive in each region can also be modeled more simply using the sub-matrix \(\mathbf{O}_{y,a}\). The general model for the vector of abundance at age in year \(y\) at age \(a\) given the vector in year \(y-1\) at age \(a-1\) is
\begin{equation*}
\log\left(\mathbf{N}_{O,y,a}\right)|\mathbf{N}_{O,y-1,a-1} =  \log\left(\mathbf{O}_{y-1,a-1}^T \mathbf{N}_{O,y-1,a-1}\right) + \boldsymbol{\varepsilon}_{y,a}
\end{equation*}
for ages \(a = 2,\ldots, A-1\), and for the plus group
\begin{equation*}
\log\left(\mathbf{N}_{O,y,A}\right)|\mathbf{N}_{O,y-1,A-1},\mathbf{N}_{O,y-1,A} = \log\left(\mathbf{O}_{y-1,A-1}^T \mathbf{N}_{O,y-1,A-1} + \mathbf{O}_{y-1,A}^T \mathbf{N}_{O,y-1,A}\right) + \boldsymbol{\varepsilon}_{y,A},
\end{equation*}
where \(\boldsymbol{\varepsilon}_{y,a}\) is the vector of region-specific errors for a given stock, which are independent across stocks and regions, but the same correlation structures as previous versions are possible across ages and years for a given stock and region.

@@

Initial numbers at age for each stock and region can be treated as age-specific fixed effects or with an equilibrium assumption as in previous versions of WHAM. For the equilibrium option there are two parameters for each stock: the stock-specific equilibrium fully-selected fishing mortality rate (\(\widetilde{F}_1\)) and the recruitment in year 1 (\(N_{1,1,r_s}\)). A stock-specific equilibrium fishing mortality at age by fleet (\(\widetilde{F}_{f,1,a}\)) is the product of \(\widetilde{F}_1\) and the selectivity across fleets in the first year
\begin{equation*}
  \text{sel}_{f,1,a} = \frac{F_{f,1,a}}{\max_a \sum_{f=1}^{n_F} F_{f,1,a}}.
\end{equation*}
We use \(\widetilde{F}_{f,1,a}\) to define an equilibrium abundance per recruit by region at age \(a\) conditional on recruiting to each region
\begin{equation}\label{eq:npraa}
 \widetilde{\mathbf{O}}_{a} = \left\{
 \begin{array}{ll}
\prod^{a-1}_{j=0}\mathbf{O}_{j}  & 1\leq a<A\\
\left[\prod^{a-1}_{j=0}\mathbf{O}_{j}\right] \left(\mathbf{I} - \mathbf{O}_{A}\right)^{-1} & a = A
 \end{array},
\right.
\end{equation}
where \(\mathbf{O}_{j}\) is the equilibrium proportions surviving age \(a\) and occurring in each region and \(\mathbf{O}_{0} = \mathbf{I}\), the identity matrix. Natural mortality and movement rates in the first year of the model are also used in Eq. \ref{eq:npraa}. For the plus group \(a=A\), \(\left(\mathbf{I} - \mathbf{O}_{A}\right)^{-1}\) is a ``fundamental matrix'\,' derived using the matrix version of the geometric series (Kemeny and Snell 1960). Recall that recruitment for a given stock \(s\) only occurs in region \(r_s\) so, the equilibrium initial numbers at age \(a\) by region are
\[\mathbf{N}_{O,1,a} = \widetilde{\mathbf{O}}_{a,r_s,\cdot} N_{1,1,r_s},\]
where \(\widetilde{\mathbf{O}}_{a,r_s,\cdot}\) is the vector of values in row \(r_s\) of \(\widetilde{\mathbf{O}}_{a}\). Initial abundance at age can also be estimated as random effects as described in Appendix B.

@@

When treated as constant parameters, natural mortality rates may be stock-, region-, and age-specific, but currently are constant across seasons. When random effects are used, the same 2DAR1 structure with age and year as described by Stock and Miller (2021) can be configured for a given stock and region and any environmental covariate effects can be stock-, region-, and age-specific. We provide details in Appendix B.

@@

The log-normal distributional assumption for aggregate catch observations is the same as Stock and Miller (2021), but the predicted catch is now a function of catch from each stock starting the year in each region and any seasonal movement. For a given stock \(s\) at age \(a\), the vector of numbers captured by each fleet over the year \(y\) is
\[\mathbf{N}_{H,s,y,a} = \mathbf{H}_{s,y,a}^T \mathbf{N}_{O,s,y,a}\]
(recall each fleet can harvest in only one region as described above). The vector of numbers caught by each fleet across stocks is
\[\mathbf{N}_{H,y,a} = \sum^{n_S}_{s=1} \mathbf{N}_{H,s,y,a}\]
and the vector of predicted aggregate catch by each fleet at age \(a\) is
\[\widehat{\mathbf{C}}_{y,a} = \text{diag}\left(\mathbf{c}_{y,a}\right) \mathbf{N}_{H,y,a},\]
where \(\mathbf{c}_{y,a}\) is the vector of mean individual weight at age \(a\) for each fleet and the aggregate catch by fleet is
\[\widehat{\mathbf{C}}_y = \sum^{A}_{a=1} \widehat{\mathbf{C}}_{y,a}.\]
The log-aggregate catch observations for fleet \(f\) are normally distributed
\[ \log C_{f,y} \sim \text{N}\left(\log \widehat {C}_{f,y}, \sigma^2_{f,y}\right),\]

where \(\widehat C_{f,y}\) is an element of \(\widehat{\mathbf{C}}_y\). The predicted numbers caught for each fleet \(f\) (row \(f\) of \(\mathbf{N}_{H,y,a}\)) are used to make predicted age composition observations as described by Stock and Miller (2021). Since then, three additional likelihood options for age composition observations have been added: a logistic-normal with AR(1) correlation structure (Francis 2014), the alternative Dirichlet-multinomial parameterization described by Thorson et al. (2017), and the multivariate Tweedie (Thorson et al. 2023).

@@

BSB are a temperate reef fish in the western Atlantic Ocean ranging the entire east coast of the United States. Fish north of Cape Hatteras, NC, are considered to comprise a single NEUS stock unit, but individual populations exhibit spawning site fidelity (Able and Stanton Hales, Jr. 1997; Fabrizio et al. 2013). Fish in this NEUS stock perform seasonal migrations out on the continental shelf in the fall and back to their inshore spawning areas during the spring (Musick and Mercer 1977). Analyses of tagging studies have demonstrated that the extent of seasonal migration varies along the coast such that fish from populations off of southern New England and further north move offshore and as far south as the coasts of Virgina and North Carolina, whereas fish in the southern portion of the stock range generally move shorter distances between inshore and offshore areas of similar latitude (Moser and Shepherd 2009).

>>>>

\externaldocument{supp_accepted} %allows references from supp_accepted.aux

@@

State-space age-structured stock assessment models can be used to estimate time and age-varying population attributes as random effects using maximum marginal likelihood or Bayesian fitting procedures (\protect\hyperlink{ref-nielsenberg14}{Nielsen and Berg 2014}; \protect\hyperlink{ref-cadigan16}{Cadigan 2016}; \protect\hyperlink{ref-milleretal16}{Miller et al. 2016b}). This estimation approach is considered an essential feature of gold-standard assessment models that we use in tactical management of commercially important fish stocks (\protect\hyperlink{ref-puntetal20}{Punt et al. 2020}). The State-space Assessment Model (SAM, \protect\hyperlink{ref-nielsenberg14}{Nielsen and Berg 2014}) continues to be developed and remains widely used within the International Council for the Exploration of the Sea (ICES) to assess European fish stocks. Various state-space models are being used to manage cod and plaice stocks in the waters of Eastern Canada (\protect\hyperlink{ref-perreaultetal20}{Perreault et al. 2020}; \protect\hyperlink{ref-varkeyetal22}{Varkey et al. 2022}) and to the south, the Woods Hole Assessment Model (WHAM, \protect\hyperlink{ref-stockmiller21}{Stock and Miller 2021}) is now used to assess many fish stocks in the Northwest Atlantic Ocean (\protect\hyperlink{ref-nefsc22}{NEFSC 2022a}, \protect\hyperlink{ref-nefsc22a}{2022b}; \protect\hyperlink{ref-nefsc24}{NEFSC 2024a}).

WHAM is an R package developed and maintained by scientists at NOAA's Northeast Fisheries Science Center (\url{https://timjmiller.github.io/wham}). WHAM can be configured to fit a wide range of age-structured models from traditional statistical catch-at-age models without any random effects, to models with several time and age varying process errors and effects of environmental covariates on various population parameters. Like SAM, WHAM models are built using the Template Model Builder package (TMB, \protect\hyperlink{ref-kristensenetal16}{Kristensen et al. 2016}) which provides a computationally efficient means of fitting an extremely wide class of models with random effects. WHAM has undergone active development since its creation and includes random effects options for recruitment, inter-annual transitions of numbers at age (hereafter referred to as ``apparent survival''), fishery and index selectivity, natural mortality, and catchability.

However, WHAM has only allowed consideration of a single stock and region, and no seasonal changes in stock dynamics. Using such models for stocks that have subcomponents with varying seasonal movement can provide incorrect inferences and poor management advice (\protect\hyperlink{ref-yingetal11}{Ying et al. 2011}; \protect\hyperlink{ref-caoetal14}{Cao et al. 2014}; \protect\hyperlink{ref-bosleyetal22}{Bosley et al. 2022}). Furthermore, the ability to account for spatial structure and model multiple stocks are also important features of leading-edge assessment modeling frameworks (\protect\hyperlink{ref-puntetal20}{Punt et al. 2020}). We describe here the implementation of these features and other extensions since Stock and Miller (\protect\hyperlink{ref-stockmiller21}{2021}) in WHAM version 2.0. Many of these new configuration options can also be useful when modeling a single stock and region.

We developed this extension of WHAM in concert with a research track assessment process for black sea bass (BSB, \textit{Centropristis striata}; \protect\hyperlink{ref-nefsc23}{NEFSC 2023}), which is a high profile stock in the Northeast U.S. and important to both commercial fishing fleets and recreational anglers. BSB are targeted from Maine to North Carolina in both state and federal waters. Past stock assessments have estimated that the stock is healthy (not overfished and overfishing not occurring). Those assessments assumed two stock components (north and south), each assessed with a separate statistical catch at age model, with results subsequently combined to inform management. However, the assessments exhibited strong retrospective patterns. The stock was split into two regional components to account for spatial dynamics and improve model diagnostics (\protect\hyperlink{ref-asmfc16}{ASMFC 2016}). Additionally, BSB have complex spatial dynamics and make seasonal migrations, moving inshore in the spring and offshore in the fall (\protect\hyperlink{ref-mosershepherd09}{Moser and Shepherd 2009}). BSB distribution has been linked to warming waters on the northwest shelf (\protect\hyperlink{ref-belletal15}{Bell et al. 2015}) and it is hypothesized that the distribution and productivity of the BSB stock is especially susceptible to long-term changes in temperature causing increases in abundance further north (\protect\hyperlink{ref-hareetal16}{Hare et al. 2016}). Thus, a focus of the BSB research track assessment was to develop an assessment platform that could capture multiple stock components, complex seasonal movement, and explore environmental drivers of population dynamics (\protect\hyperlink{ref-nefsc23}{NEFSC 2023}). We denote the fish in the north as separate stock components rather than stocks because both components are managed as a single stock, but recognizing differences in migration between fish originating in the two areas.

To illustrate the usage of WHAM 2.0 we apply it to the two stock components of BSB off the coast of the Northeast United States (NEUS) using most of the model assumptions that were accepted during the peer-review process (\protect\hyperlink{ref-nefsc23}{NEFSC 2023}). However, here we evaluate evidence for alternative hypotheses of temporal variation and effects of bottom temperature, specifically, on recruitment and natural mortality of age 1 individuals.

When modeling multiple stock components and regions, further dimensionality is added to the class of possible equilbrium reference points (e.g., \protect\hyperlink{ref-kapuretal21}{Kapur et al. 2021}). Short term projections are also an important part of the management process and the separation of observation and process errors in state-space models allows uncertainty in projected attributes such as recruitment and spawning stock biomass (SSB) without external simulation exercises. Finally, the inclusion of explicit effects of environmental covariates in the model also requires assumptions on how they are treated in projections and reference points. We demonstrate some of the options of WHAM version 2.0 to provide biological reference points and short-term projections.

@@

Many of the options and equations of WHAM version 2.0 are the same as those in Stock and Miller (\protect\hyperlink{ref-stockmiller21}{2021}), so we will only describe extensions and differences that have occurred since their first description of WHAM. The new version of WHAM can model multiple stocks, each with their own movement, harvest, and natural mortality. Much of the description below is for a specific stock \(s\), but, for simplicity, this subscript is implicit except when necessary. We have provided an Appendix A Table \ref{symbols} containing definitions of all mathematical notation in this paper.

@@

Because individuals may be alive in one of several regions or harvested by one of several fleets, it is helpful to consider these as distinct categories or states and treat the number of individuals occurring in each category over time as a multi-state model. Approaches to modeling transitions among these states may treat time discretely (e.g., \protect\hyperlink{ref-arnason72}{Arnason 1972}; \protect\hyperlink{ref-schwarzetal93}{Schwarz et al. 1993}) or continuously (e.g., \protect\hyperlink{ref-hearnetal87}{Hearn et al. 1987}; \protect\hyperlink{ref-commenges99}{Commenges 1999}; \protect\hyperlink{ref-andersenkeiding02}{Andersen and Keiding 2002}). We can define a probability transition matrix (PTM) that describes the probability of individuals occurring in different states at the end of a time interval with duration \(\delta\) (columns), conditional on being in each of those states at the beginning of the interval (rows). For fish populations, these states would be defined as being alive in a particular region or deceased due to fishing from a particular fleet or natural mortality. The time interval \(i\) with duration \(\delta_i\) would be a season and the PTMs would be uniquely defined for each stock by year \(y\), age \(a\), and season \(i\). Each row and column of the PTM correspond to one of the states: alive in region \(r\), dead in fleet \(f\), or dead from natural causes. The probabilities in each row sum to one (must be in one of the states) and assume an individual is in the state that corresponds to that row at the beginning of the interval. Given \(n_R\) regions and \(n_F\) fleets, the square PTM (\(n_R + n_F + 1\) rows and columns) as a function of sub-matrices is
\begin{equation}\label{eq:ptm}
  \mathbf{P}_{y,a,i} = \begin{bmatrix}
    \mathbf{O}_{y,a,i} & \mathbf{H}_{y,a,i} & \mathbf{D}_{y,a,i} \\
    0 & \mathbf{I}_{H} & 0\\
    0 & 0 & 1
  \end{bmatrix},
\end{equation}
where
\begin{equation*}
  \mathbf{O}_{y,a,i} = 
  \begin{bmatrix}
    O_{y,a,i}(1,1) & \cdots & O_{y,a,i}(1,n_R) \\
    \vdots & \ddots & \vdots \\
    O_{y,a,i}(n_R,1) & \cdots & O_{y,a,i}(n_R,n_R)
  \end{bmatrix}
\end{equation*}
is the \(n_R \times n_R\) matrix defining survival and occurring in each region at the end of the interval,
\begin{equation*} 
  \mathbf{H}_{y,a,i} = 
  \begin{bmatrix}
    H_{y,a,i}(1,1) & \cdots & H_{y,a,i}(1,n_F) \\
    \vdots & \ddots & \vdots \\
    H_{y,a,i}(n_R,1) & \cdots & H_{y,a,i}(n_R,n_F)
  \end{bmatrix}
\end{equation*}
is the \(n_R \times n_F\) matrix defining probabilities of being captured in each fleet during the interval, and \(\mathbf{D}_{y,a,i}\) is the \(n_R \times 1\) matrix of probabilities of dying due to natural mortality during the interval. We have the \(n_F\times n_F\) identity matrix \(\mathbf{I}_{H}\) for the states of capture by each fleet and a 1 for the natural mortality state because the probabilities of being in one of the mortality states given starting the interval in that state is one (no zombies allowed).

@@

When there is more than one region, WHAM can model survival and movement as processes occurring sequentially or simultaneously. The sequential assumption is used widely in spatially explicit models (e.g., Stock Synthesis, \protect\hyperlink{ref-methotwetzel13}{Methot and Wetzel 2013}). Under the sequential assumption, survival and death occur over the interval and movement among regions occurs instantly at either the beginning or the end of the interval. WHAM is configured to have movement occur after survival:
\begin{equation*}
  \mathbf{O}_{y,a,i} = \mathbf{S}_{y,a,i}\boldsymbol{\mu}_{y,a,i},
\end{equation*}
where \(\mathbf{S}_{y,a,i}\) is a \(n_R \times n_R\) diagonal matrix of proportions surviving in each region (given they start in that region)
\begin{equation*}
\mathbf{S}_{y,a,i} = 
  \begin{bmatrix}
    e^{-Z_{y,a,i,1}\delta_i} & 0 & \cdots & 0 \\
    0 & e^{-Z_{y,a,i,2}\delta_i} & \cdots & 0 \\
    \vdots & \vdots & \ddots & \vdots \\
    0 & \cdots & 0 & e^{-Z_{y,a,i,n_R}\delta_i}
  \end{bmatrix}
\end{equation*}
and \(\boldsymbol{\mu}_{y,a,i}\) is a \(n_R \times n_R\) matrix of proportions moving from one region to another or staying in the region they occurred at the beginning of the interval:
\begin{equation*}
\boldsymbol{\mu}_{y,a,i} = 
  \begin{bmatrix}
    1-\sum_{r' \neq 1} \mu_{1\rightarrow r',y,a,i} & \mu_{1\rightarrow 2,y,a,i} & \cdots & \mu_{1\rightarrow R,y,a,i} \\
    \mu_{2\rightarrow 1,y,a,i} & 1-\sum_{r' \neq 2} \mu_{2\rightarrow r',y,a,i} & \cdots & \mu_{2\rightarrow R,y,a,i} \\
    \vdots & \vdots & \ddots & \vdots \\
    \mu_{R\rightarrow 1,y,a,i} & \cdots & \mu_{R\rightarrow R-1,y,a,i} & 1-\sum_{r' \neq R} \mu_{R\rightarrow r',y,a,i}
  \end{bmatrix}
\end{equation*}

@@

As in previous versions of WHAM, the apparent survival (and movement) from one year to another after recruitment can be treated deterministically or as functions of random effects. The predicted numbers at age \(a\) in year \(y\) for a given stock are vector analogs (\(\mathbf{N}_{O,y,a}\)) of the equations for numbers at age in the standard WHAM model (\protect\hyperlink{ref-stockmiller21}{Stock and Miller 2021}). For ages \(a = 2,\ldots, A-1\), where \(A\) is the plus group, the expected number alive in each region at the beginning of the following year and next age class age can be obtained from the first \(n_R\) elements of the vector
\[\mathbf{P}_{y-1,a-1}^T \mathbf{N}_{y-1,a-1}.\]
The numbers alive in each region can also be modeled more simply using the sub-matrix \(\mathbf{O}_{y,a}\). The general model for the vector of abundance at age in year \(y\) at age \(a\) given the vector in year \(y-1\) at age \(a-1\) is
\begin{equation*}
\log\left(\mathbf{N}_{O,y,a}\right)|\mathbf{N}_{O,y-1,a-1} =  \log\left(\mathbf{O}_{y-1,a-1}^T \mathbf{N}_{O,y-1,a-1}\right) + \boldsymbol{\varepsilon}_{y,a}
\end{equation*}
for ages \(a = 2,\ldots, A-1\), and for the plus group
\begin{equation*}
\log\left(\mathbf{N}_{O,y,A}\right)|\mathbf{N}_{O,y-1,A-1},\mathbf{N}_{O,y-1,A} = \log\left(\mathbf{O}_{y-1,A-1}^T \mathbf{N}_{O,y-1,A-1} + \mathbf{O}_{y-1,A}^T \mathbf{N}_{O,y-1,A}\right) + \boldsymbol{\varepsilon}_{y,A},
\end{equation*}
where \(\boldsymbol{\varepsilon}_{y,a}\) is the vector of region-specific errors for a given stock, which are independent across stocks and regions, but the same correlation structures as previous versions are possible across ages and years for a given stock and region.

@@

Initial numbers at age for each stock and region can be treated as age-specific fixed effects or with an equilibrium assumption as in previous versions of WHAM. For the equilibrium option there are two parameters for each stock: the stock-specific equilibrium fully-selected fishing mortality rate (\(\widetilde{F}_1\)) and the recruitment in year 1 (\(N_{1,1,r_s}\)). A stock-specific equilibrium fishing mortality at age by fleet (\(\widetilde{F}_{f,1,a}\)) is the product of \(\widetilde{F}_1\) and the selectivity across fleets in the first year
\begin{equation*}
  \text{sel}_{f,1,a} = \frac{F_{f,1,a}}{\max_a \sum_{f=1}^{n_F} F_{f,1,a}}.
\end{equation*}
We use \(\widetilde{F}_{f,1,a}\) to define an equilibrium abundance per recruit by region at age \(a\) conditional on recruiting to each region
\begin{equation}\label{eq:npraa}
 \widetilde{\mathbf{O}}_{a} = \left\{
 \begin{array}{ll}
\prod^{a-1}_{j=0}\mathbf{O}_{j}  & 1\leq a<A\\
\left[\prod^{a-1}_{j=0}\mathbf{O}_{j}\right] \left(\mathbf{I} - \mathbf{O}_{A}\right)^{-1} & a = A
 \end{array},
\right.
\end{equation}
where \(\mathbf{O}_{j}\) is the equilibrium proportions surviving age \(a\) and occurring in each region and \(\mathbf{O}_{0} = \mathbf{I}\), the identity matrix. Natural mortality and movement rates in the first year of the model are also used in Eq. \ref{eq:npraa}. For the plus group \(a=A\), \(\left(\mathbf{I} - \mathbf{O}_{A}\right)^{-1}\) is a ``fundamental matrix'\,' derived using the matrix version of the geometric series (\protect\hyperlink{ref-kemenysnell60}{Kemeny and Snell 1960}). Recall that recruitment for a given stock \(s\) only occurs in region \(r_s\) so, the equilibrium initial numbers at age \(a\) by region are
\[\mathbf{N}_{O,1,a} = \widetilde{\mathbf{O}}_{a,r_s,\cdot} N_{1,1,r_s},\]
where \(\widetilde{\mathbf{O}}_{a,r_s,\cdot}\) is the vector of values in row \(r_s\) of \(\widetilde{\mathbf{O}}_{a}\). Initial abundance at age can also be estimated as random effects as described in Appendix B.

@@

When treated as constant parameters, natural mortality rates may be stock-, region-, and age-specific, but currently are constant across seasons. When random effects are used, the same 2DAR1 structure with age and year as described by Stock and Miller (\protect\hyperlink{ref-stockmiller21}{2021}) can be configured for a given stock and region and any environmental covariate effects can be stock-, region-, and age-specific. We provide details in Appendix B.

@@

The log-normal distributional assumption for aggregate catch observations is the same as Stock and Miller (\protect\hyperlink{ref-stockmiller21}{2021}), but the predicted catch is now a function of catch from each stock starting the year in each region and any seasonal movement. For a given stock \(s\) at age \(a\), the vector of numbers captured by each fleet over the year \(y\) is
\[\mathbf{N}_{H,s,y,a} = \mathbf{H}_{s,y,a}^T \mathbf{N}_{O,s,y,a}\]
(recall each fleet can harvest in only one region as described above). The vector of numbers caught by each fleet across stocks is
\[\mathbf{N}_{H,y,a} = \sum^{n_S}_{s=1} \mathbf{N}_{H,s,y,a}\]
and the vector of predicted aggregate catch by each fleet at age \(a\) is
\[\widehat{\mathbf{C}}_{y,a} = \text{diag}\left(\mathbf{c}_{y,a}\right) \mathbf{N}_{H,y,a},\]
where \(\mathbf{c}_{y,a}\) is the vector of mean individual weight at age \(a\) for each fleet and the aggregate catch by fleet is
\[\widehat{\mathbf{C}}_y = \sum^{A}_{a=1} \widehat{\mathbf{C}}_{y,a}.\]
The log-aggregate catch observations for fleet \(f\) are normally distributed
\[ \log C_{f,y} \sim \text{N}\left(\log \widehat {C}_{f,y}, \sigma^2_{f,y}\right),\]

where \(\widehat C_{f,y}\) is an element of \(\widehat{\mathbf{C}}_y\). The predicted numbers caught for each fleet \(f\) (row \(f\) of \(\mathbf{N}_{H,y,a}\)) are used to make predicted age composition observations as described by Stock and Miller (\protect\hyperlink{ref-stockmiller21}{2021}). Since then, three additional likelihood options for age composition observations have been added: a logistic-normal with AR(1) correlation structure (\protect\hyperlink{ref-francis14}{Francis 2014}), the alternative Dirichlet-multinomial parameterization described by Thorson et al. (\protect\hyperlink{ref-thorsonetal17}{2017}), and the multivariate Tweedie (\protect\hyperlink{ref-thorsonetal23}{Thorson et al. 2023}).

@@

BSB are a temperate reef fish in the western Atlantic Ocean ranging the entire east coast of the United States. Fish north of Cape Hatteras, NC, are considered to comprise a single NEUS stock unit, but individual populations exhibit spawning site fidelity (\protect\hyperlink{ref-ablehales97}{Able and Stanton Hales, Jr. 1997}; \protect\hyperlink{ref-fabrizioetal13}{Fabrizio et al. 2013}). Fish in this NEUS stock perform seasonal migrations out on the continental shelf in the fall and back to their inshore spawning areas during the spring (\protect\hyperlink{ref-musickmercer77}{Musick and Mercer 1977}). Analyses of tagging studies have demonstrated that the extent of seasonal migration varies along the coast such that fish from populations off of southern New England and further north move offshore and as far south as the coasts of Virgina and North Carolina, whereas fish in the southern portion of the stock range generally move shorter distances between inshore and offshore areas of similar latitude (\protect\hyperlink{ref-mosershepherd09}{Moser and Shepherd 2009}).
